---
title: "Simultaneous count model vignette"
author: "G.E. Ryan"
date: '2019-10-25'
output:
  html_document:
    df_print: paged
---

This is a vignette illustrating the "simultaneous count model" approach from [*Simultaneous‐count models to estimate abundance from counts of unmarked individuals with imperfect detection*, Conservation Biology 33(3) 697—708, G.E. Ryan et al. 2019.]( https://doi.org/10.1111/cobi.13261)

## Packages
```{r packages, result='hide'}
library(dplyr)
library(rjags)
library(R2jags)
library(jagstools)
library(ggplot2)
library(bayesplot)
```
If necessary install `jagstools` from [github](https://github.com/johnbaums/jagstools) with:

`library(remotes)`

`remotes::install_github("johnbaums/jagstools")`


## Data
Create a dataset simulating population of 1000 individuals,

A survey effort at:
 - 5 sites,
 - visited 4 times/year,
 - for 5 years
 
Around half of the population is ever observed at each site in a single sampling occasion (visit).

Probabilities of detection are constant at each site over time.

### Simulated data
```{r dat}

# population size
population.size <- 1000

# site probabilities of detection [1:5], and probability of non-detection [6]
probs <- c(0.01, 0.15, 0.04, 0.22, 0.08, 0.5)

dat <- array(
  data = rmultinom(
    n = 20, # 4 visits x 5 years
    size = population.size,
    prob = probs 
  ),
  dim = c(6,4,5), # sites +1, visits/year, years
  dimnames = list(
    "site" = sprintf("site%s", 1:6),
    "visit" = sprintf("visit%s", 1:4),
    "year" = sprintf("year%s", 1:5)
  )
)
```
Let's look at the first two years of data:
```{r dat[,,1:2]}
dat[,,1:2]
```
We can see that this simulates six sites, because while we visit five sites, this simulated sixth site represents the population that we have not observed,

### Model data
To model the observation process, we won't observe the inddividuals at site six, so we record these in the array as `NA`. We'll use `y` for these data simulated observation data to be consistent with the notation in Ryan *et al*.
```{r y}
y <- dat

y[6,,] <- NA
```
Let's have a look at the first two years of these data
```{r y[,,1:2]}
y[,,1:2]
```



## Modelling

### Write our JAGS model.
This model
For explanation of ones trick see: https://www.mrc-bsu.cam.ac.uk/wp-content/uploads/manual14.pdf
```{r jags.mod}
cat('
  data{
    for (j in 1:nvisits){
      for (k in 1:nyears){
        ones[j,k] <- 1
        # constants for ones trick
      }
    }
    C <- 10000
    # constant for ones trick
  }
  model{
    lambda ~ dgamma(1e-5, 1e-5)T(1,5000)
    # Hyper-prior for population size, truncated at 5000 to improve convergence
    
    N ~ dpois(lambda)
    # Prior estimate of population size
    
    pi[1:(nsites + 1)] ~ ddirich(alpha[1:(nsites + 1)])
    # Prior for our observation process
    
    for (k in 1:nyears){
      for (j in 1:nvisits){
        
        seen[j,k] <- sum(y[1:nsites,j,k]) # Number of individuals observed
        y[(nsites + 1),j,k] <- N - seen[j,k] # complete our array y
        
        # Likelihood written in full
        # because we cant simply do y ~ multinomial with missing data in JAGS
        for (i in 1:(nsites + 1)){
          lp[i,j,k] <- y[i,j,k]*log(pi[i]) - logfact(y[i,j,k])
        }
        lnL[j,k] <- sum(lp[1:(nsites + 1),j,k]) + logfact(N)
        log(P[j,k]) <- lnL[j,k] - log(C)
        ones[j,k] ~ dbern(P[j,k]) # ones trick
      }
    }
  }
  '
    , file = (jags.mod <- tempfile())) # This writes a temporary file with the model and assigns the path at jags.mod
```

### Prepare data for JAGS
```{r jags.data}
nsites   <- dim(y)[1] - 1 # should be 5
nvisits  <- dim(y)[2]     # should be 4
nyears   <- dim(y)[3]     # should be 5
  
alpha <- rep(1, (nsites + 1))
  
jags.params <- c("N", "pi", "lambda")

jags.inits <- function(){
  list(
    lambda = 2*sum(y[,1,1], na.rm = TRUE) # initial value to search from is 5x the total observed in the first sampling occasion
  )
}

jags.data <- list(
  "y",
  "nsites",
  "nvisits",
  "nyears",
  "alpha"
)

assign("y",       y,       envir=globalenv())
assign("alpha",   alpha,   envir=globalenv())
assign("nsites",  nsites,  envir=globalenv())
assign("nvisits", nvisits, envir=globalenv())
assign("nyears",  nyears,  envir=globalenv())
```

### Fit our model
This is a short-run test fit, with just 5000 iterations.
```{r jags.fit, results='hide'}
jags.fit <- jags(
  data = jags.data,
  inits = jags.inits,
  parameters.to.save = jags.params,
  n.iter = 5000,
  n.burnin = 100,
  n.chains = 3,
  model.file = jags.mod
)
```

```{r print jags.fit, echo=FALSE}
jags.fit
```
The high R-hat values suggest the model is not converged.

Let's also check the traceplots.
```{r trace}
color_scheme_set(scheme = "viridis")
mcmc_trace(x = jags.fit$BUGSoutput$sims.array)
```
These are clearly not converged.

Let's try fitting with a much longer run and burn-in. Generally this might be done iteratively, but let's go straight to 300k iterations with 100k burn-in.

We also use `jags.parallel` here to fit chains in parallel.
```{r jags.fit parallel, results='hide'}
jags.fit <- do.call(
  what = jags.parallel,
  args = list(
    data = jags.data,
  inits = jags.inits,
  parameters.to.save = jags.params,
  n.iter = 300000,
  n.burnin = 100000,
  n.chains = 3,
  n.thin = 100,
  model.file = jags.mod
  )
)
```

```{r print jags.fit parallel, echo=FALSE}
jags.fit
```
These R-hat values are much better (<1.1).

Let's also see some traceplots.
```{r trace parallel}
mcmc_trace(x = jags.fit$BUGSoutput$sims.array)
```
Showing a lot of autocorrelation, but appear to be converged.

## Results

Let's have a look at our results.

Yellow points represent the posterior median, dark purple bands the 50% credible interval, and light blue bands the 95% credible interval.

Our estimates of *pi*:
```{r mcmc_intervals pi}
mcmc_intervals(
  x = jags.fit$BUGSoutput$sims.array,
  pars = sprintf("pi[%s]", 1:6),
  prob_outer = 0.95
)
```

And of *N*:
```{r mcmc_intervals N}
mcmc_intervals(
  x =jags.fit$BUGSoutput$sims.array,
  pars = "N",
  prob_outer = 0.95
) +
  scale_x_continuous(limits = c(0, NA))
```

And let's compare these directly with our true values. Here the red box and cross represent our true value, plotted over our posterior estimates.

*Pi*:
```{r mcmc_intervals pi true}
true.pi <- data.frame(
  x = probs,
  y = sprintf("pi[%s]", 1:6)
)

mcmc_intervals(
  x = jags.fit$BUGSoutput$sims.array,
  pars = sprintf("pi[%s]", 1:6),
  prob_outer = 0.95
) +
  geom_point(
    data = true.pi,
    aes(
      x = x,
      y = y
    ),
    shape = 7,
    col = "red",
    size = 5
  )
```
Lookds pretty good!

And of *N*:
```{r mcmc_intervals N true}
true.N <- data.frame(
  x = population.size,
  y = "N"
)

mcmc_intervals(
  x =jags.fit$BUGSoutput$sims.array,
  pars = "N",
  prob_outer = 0.95
) +
  scale_x_continuous(limits = c(0, NA)) +
  geom_point(
    data = true.N,
    aes(
      x = x,
      y = y
    ),
    shape = 7,
    col = "red",
    size = 5
  )
```
Not too bad.